---
title: "Raji IVIS"
author: "L. Meuris"
date: "2025-12-02"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
here::i_am("analysis/Raji_IVIS.Rmd")
library(here)
library(readxl)
library(tidyverse)
library(scales)

power_trans <- function(power){
   trans_new("power",
             transform = function(y){
                if(power != 0){
                   yt <- sign(y) * abs(y)^(power)
                } else {
                   yt = y
                }
                return(yt)
             },
             inverse = function(yt){
                if(power != 0){
                   y <- yt^(1/(power))
                } else {
                   y <- yt
                   
                }
                return(y)
             },
             domain = c(-Inf, Inf)
             )
}
```

# Description of experiments

Purpose of the experiment is to look at the effect of MGAT5 engineering in CART-cells in a mouse tumor model. The model uses Raji cells, which create blood tumors of the Burkitt's lymphoma type. These cells are CD19, CD20 and CD70 positive.  

The experiment has the following timeline:

  - day 0: injection of 500 000 Raji cells IV
  - around day 6: IVIS baseline measurement
  - around day 7: treatment (CAR T injection ivin different doses and controls)
  - around day 133: rechallenge in some targets
  
The mice are followed up via body weight measurements and IVIS data, which measures luminescence of the Raji cells in the mice. IVIS measurements are taken at set days (every 3 or 4 days) and mice weights are recorded every 2 days. There are 4 experiments, each with 5 mice per group. 

Experiment 51:

  - PBS: no cells given
  - NTC: 4 million cells donor T-cells, no CAR T cells
  - CD70 CART WT 4 million cells
  - CD70 CART MGAT5 4 million cells
  - CD70 CART WT 2 million cells
  - CD70 CART MGAT5 2 million cells
  - CD20 CART WT 1 million cells
  - CD20 CART MGAT5 1 million cells
  - CD20 CART WT 0.5 million cells
  - CD20 CART MGAT5 0.5 million cells
  
Experiment 62 and 68:

  - NTC: 4 million cells donor T-cells, no CAR T cells
  - CD70 CART WT 4 million cells
  - CD70 CART MGAT5 4 million cells
  - CD20 CART WT 0.5 million cells
  - CD20 CART MGAT5 0.5 million cells
  - CD19 CART WT 2 million cells
  - CD19 CART MGAT5 2 million cells
  
Experiment 74: used 6 mice per group, and no CD19 CART cells:

  - NTC: 4 million cells donor T-cells, no CAR T cells
  - CD70 CART WT 4 million cells
  - CD70 CART MGAT5 4 million cells
  - CD20 CART WT 0.5 million cells
  - CD20 CART MGAT5 0.5 million cells
  
Experiment 91 and 97

  - NTC 
  - CD19 CART WT
  - CD19 CART MGAT5
  
Summarizing we can say that: 

  1) there are 3 CART targets: CD19, CD20, CD70
  2) different amounts of CART cells have been injected; 0.5, 2 or 4 million
  3) CART cells were either engineered or not to have MGAT5 knocked out
  
There was a rechallenge part to experiments 51 and 74, but not to 62 and 68. No data yet on exp 91 or 97, experiments still running. A rechallenge means that tumor cells were administered again after a set period.
We also note the end-day of the experiments, as we need this for calculations of the survival times.
Experiment 91 and 97 are still running, so no end days yet there either. 

```{r}
day_rechallenge_E51 <- 133
day_rechallenge_E62 <- NA
day_rechallenge_E68 <- NA
day_rechallenge_E74 <- 139
day_rechallenge_E91 <- NA
day_rechallenge_E97 <- NA

day_end_E51 <- 215
day_end_E62 <- 83
day_end_E68 <- 88
day_end_E74 <- 202
day_end_E91 <- NA
day_end_E97 <- NA
```

# Data

First, we load all the data for Exp 51, 62, 68, 74, 91, 97
```{r}
sheets <- excel_sheets(here('data/data_cleaned.xlsx'))
datalist <- map(sheets, read_xlsx, path = here('data/data_cleaned.xlsx'))
names(datalist) <- sheets
```

## Metadata 
Get the metadata
```{r}
metadata <- datalist$Metadata
```

## IVIS data
Get the IVIS data for each experiment,
Put in long format,
Combine data from 4 experiments into one large dataframe,
Remove lines with missing weight data
Select relevant columns and sort

```{r}
IVIS <- datalist[str_detect(names(datalist), 'IVIS')] %>% # select only IVIS data
  bind_rows(.id = 'Experiment') %>%
  mutate(Experiment = str_remove(Experiment, 'IVIS'),
         Mouse = as.numeric(Mouse)) %>%
  filter(!is.na(IVIS_Counts)) %>%
  filter(IVIS_Exposure == 3) %>%
  select(Experiment, Mouse, Day, IVIS_Counts) %>%
  arrange(Experiment, Mouse, Day)
IVIS <- left_join(IVIS, metadata)
```

Clean up
```{r}
rm(metadata, datalist, sheets)
```

# Data cleaning

Parse variables to correct type
Remove some data that are superfluous or problematic: 

EXP51:
  - Mouse 21 in exp 51 was a sentinel mouse

EXP62:
  - Mouse 26 and 27 in exp 62 were spare, were not used
  - Mouse 18, 23, 24, 25, 28 in exp 62 died from heat of the lamps so cannot be used in analysis

EXP68:
  - Mice 1, 22 were not in the experiment and 38-43 were backup in exp 68.
  
EXP74: 
  - Mouse 22: euthanized due to unsure Raji injection
  - Mouse 30: euthanized by accident
  - Mouse 29: excluded due to unsure Raji injection
  - Mouse 46 and 47: controls for Raji effectiveness at secondary injection


```{r}
IVIS <- IVIS %>%
  filter(!(Mouse %in% c(18,23,24,25,26,27,28) & Experiment == 'E62')) %>%
  filter(!(Mouse == 21 & Experiment == 'E51')) %>%
  filter(!(Mouse %in% c(1,22,38,39,40,41,42,43) & Experiment == 'E68')) %>%
  filter(!(Mouse %in% c(22,29,30,46,47) & Experiment == 'E74')) %>%
  arrange(Experiment, Mouse, Day) %>%
  mutate(Experiment = factor(Experiment), 
         ID = fct_inorder(paste0(Experiment, 'M', Mouse)), 
         Treatment = factor(Treatment), 
         Target = factor(Target, levels =c('Ctrl', 'CD19', 'CD20', 'CD70')), 
         MGAT5 = factor(MGAT5, levels = c('Ctrl', 'MGAT5WT', 'MGAT5KO')),
         Dose = factor(Dose, levels = c('0','1','0.5','2','4'))) %>%
  select(Experiment, Mouse, ID, everything())
```


Write datasheets to xlsx files for later reference
```{r}
#data$IVIS %>% writexl::write_xlsx(here('output/IVIS.xlsx'))
```

Some data QC
```{r}
#data$IVIS %>% map(~any(is.na(.)))
```

# Plot IVIS data 

IVIS traces for individual mice per experiment

## CD70

```{r}
IVIS_70 <- IVIS %>%
  filter(!Experiment %in% c('E91', 'E97')) %>%
  filter((Target == 'CD70' & Dose == '4') | Target ==  'Ctrl') %>%
  filter(ifelse(Experiment == 'E51', Day < day_rechallenge_E51, Day < day_rechallenge_E74)) %>%
  droplevels %>%
  ggplot(aes(x = Day, y = IVIS_Counts, Group = ID, color = MGAT5)) +
  geom_line(alpha = 0.8) +
  facet_grid(Experiment~.)+
  theme_bw() +
  scale_y_log10() +
  scale_color_manual(values = c(
    rgb(44,20,83, maxColorValue = 255), 
    rgb(14,111,124, maxColorValue = 255),
    rgb(251,79,6, maxColorValue = 255))) +
  ylab('IVIS counts') +
  ggtitle('CD70 CART cells')

IVIS_70

ggsave(here('output/IVISPlot_CD70_Raji_perExp.pdf'), IVIS_70, 
       device = 'pdf', width  = 8, height = 5)
```

## CD19

```{r}
IVIS_19 <- IVIS %>%
  filter(Experiment == 'E91' | Experiment == 'E97') %>%
  droplevels %>%
  ggplot(aes(x = Day, y = IVIS_Counts, Group = ID, color = MGAT5)) +
  geom_line(alpha = 0.8) +
  facet_grid(Experiment~.)+
  theme_bw() +
  scale_y_log10() +
  scale_color_manual(values = c(
    rgb(44,20,83, maxColorValue = 255), 
    rgb(14,111,124, maxColorValue = 255),
    rgb(251,79,6, maxColorValue = 255))) +
  ylab('IVIS counts') +
  ggtitle('CD19 CART cells')

IVIS_19

ggsave(here('output/IVISPlot_CD19_Raji_perExp.pdf'), IVIS_19, 
       device = 'pdf', width  = 8, height = 5)
```

Plot per mouse to check for errors

```{r}
IVIS_19_permouse <- IVIS %>%
  filter(Experiment == 'E97') %>%
  droplevels %>%
  ggplot(aes(x = Day, y = IVIS_Counts, Group = ID, color = MGAT5)) +
  geom_line(alpha = 0.8) +
  facet_wrap(~fct_reorder(ID, as.numeric(Treatment))) +
  theme_bw() +
  scale_y_log10() +
  scale_color_manual(values = c(
    rgb(44,20,83, maxColorValue = 255), 
    rgb(14,111,124, maxColorValue = 255),
    rgb(251,79,6, maxColorValue = 255))) +
  ylab('IVIS counts') +
  ggtitle('CD19 CART cells')

IVIS_19_permouse

ggsave(here('output/IVISPlot_CD19_Raji_perMouse.pdf'), IVIS_19_permouse, 
       device = 'pdf', width  = 10, height = 10)
```

